# Study of the damages caused by storms and weather events on economy and public health in the U.S between 1950 and 2011



## Synopsis

This report aims to study the consequences of the storms and severe weather events on the 
economy and public health in the United States between 1950 and 2011.

This project involves exploring the U.S. National Oceanic and Atmospheric
Administration's (NOAA) storm database. This database tracks characteristics of
major storms and weather events in the United States, including when and where
they occur, as well as estimates of any fatalities, injuries, and property
damage.

**TODO: Complete the synopsis**  


## Data processing

### Loading and preprocessing of the raw data

We load the Storm data from a CSV file downloaded from the U.S National Oceanic
and Atmospheric Administration's (NOAA) storm database.

```{r loading,cache=TRUE}
storm <- read.csv("data/repdata-data-StormData.csv")
dim(storm)
head(storm)
```

The the Storm data contains 902297 rows and 37 columns.

We format properly the columns `BGN_DATE` and `END_DATE`:

```{r clean_dates, cache=TRUE}
bgn_dates <- as.Date(as.character(storm$BGN_DATE), "%m/%d/%Y")
end_dates <- as.Date(as.character(storm$END_DATE), "%m/%d/%Y")
```


## Results

### The most harmful events with respect to population health

In this section we focus on the consequences of the weather events on the population health. In particular we'll set a ranking of the most harmful events designated by the variable `EVTYPE`. 

Here are the 10 first most frequently recorded event types:
```{r evtypes}
head(rev(sort(table(storm$EVTYPE))), 10)
```


The variables of interest to determine which types of events are most harmful with respect to health population are the `FATALITIES` and `INJURIES`.

```{r}
library(dplyr)

injuries <- storm %>% group_by(EVTYPE) %>% summarise(total=sum(INJURIES)) %>% arrange(-total)
fatalities <- storm %>% group_by(EVTYPE) %>% summarise(total=sum(FATALITIES)) %>% arrange(-total)
```


### The types of events that had the greatest economic consequences